\thetitle

### 2020-07-02 writing/challenges.tex
working on challenges

The most serious challenge concerns the nature of this research. Unlike ordinary
work on programming languages, the question seems to call for research similar
to the one in the human-computer interaction area. At the same time, a
significant evaluation result demands a good number of cases yielding a serious
amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
around this dilemma is to postulate the equivalent of a {\it home economicus\/}
for programming languages---the {\em rational programmer\/}---and to implement
this rational programmer as an algorithm. As indicated in the introduction, in
the context of gradual typing a rationally acting programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion.

One derivative challenge to the first is the problem of simulating the
programmer's mistakes. Before a programmer can try to find mistakes, mistakes
have to be made. Technically the problem is the lack of a sufficiently large
available corpus of type errors for a meaningful experiment. Similar to
~\citet{lksfd-popl-2020}, we cover this gap by injecting bugs in typed programs
with mutation analysis~\cite{lipton1971fault, demillo1978hints,
jia2011analysis}. This results into a large number of buggy programs whose
mix-typed versions are suitable subjects for our experiment. However, mutation
analysis traditionally aims to inject bugs that are sophisticated enough to
challenge extensive test suites. As a consequence, the ill-typed mutants we care
about here are usually deemed \emph{incompetent} and mutation analysis
frameworks are fine-tuned to avoid them. To overcome this problem, we design a
set of mutators specific for type errors inspired by common errors when
programming with in Typed Racket.  These type errors cover a wide range of Typed
Racket's type system and moreover cause reliably runtime type errors in
mix-typed versions of mutants.

Another derivative challenge is the problem of mistakes in type specifications.
Gradual typing enables programmers to move an untyped code base into the typed
realm on an incremental basis. The retroactive addition of a specific type to a
component, even a library, may result from a misunderstanding of the code and
may thus be a mistake itself. Empirical evidence suggests that this scenario is
reasonably common~\cite{sta-nt-base-types, incorrect-ts, wmwz-ecoop-2017}, and
hence must be addressed by any evaluation method.  \ldots needs to also explore
the relation between blame and the location of bugs in case of wrong type
annotations. Similarly to the previous challenge, we deal with this one by
designing a set of mutators that ``break'' type annotations but leave programs
otherwise intact.

The second big challenge is the set of seemingly incomparable blame-assignment
strategies. This results in this paper focus on three points of the landscape:
the {\em natural\/} approach of Typed Racket~\cite{tf-dls-2006, tf-popl-2008,
tfffgksst-snapl-2017, tf-icfp-2010}, the {\em transient\/} one of Reticulated
Python~\cite{vsc-dls-2019,vss-popl-2017,vksb-dls-2014}, and the {\em optional\/}
typing approach of industrial systems. Each of these points comes with a
distinct blaming strategy and together they represent all proposed strategies so
far. In addition, instead of experimenting with the three approaches in
isolation, we have implemented transient and optional gradual typing as variants
of Typed Racket to seize the opportunity to compare them.  However,
implementations on top of the same platform are not enough for an
apples-to-apples comparison. The three approaches differ not only on their
blaming strategies but also on the type checks they perform. To untangle the
two, we supplement our experiment with two extra modes. The \emph{vanilla} mode
ignores blame and instead uses information from dynamic type error messages to
select how to proceed with migration. Hence it separates the effect of blame on
the experiment from that of the type checks. The \emph{control} mode randomly
selects which part of the program to migrate and serves as the null hypothesis
control for the overall methodology.

>>>>

\cd{One point we haven's justified is why empirical. The justification
seems easy in the sense that we are looking at a real language, programs,
bugs rather than models.}

\cd{I like this idea of the competent programmer. I couldn't have thought
about homo economicus. Instead this was partly inspired by Gonen's work.}
\cd{There they make a hypothesis, try to test it in vivo using designed model organisms
with a well-specified first-order characteristic (e.g, a deleted gene).
Then they collect data by placing the organism in a number of scenarios 
and measuring aspects of its behavior and then they 
construct an operational model from their data. Due to the synthetic
nature of CS, for us the process goes in the opposite direction. In that
sense the competent programmer is a model organism that, as its
first-order characteristic, implements a direct consequence of the
semantics. So in a sense for me this has always been about semantics and
exploring their consequences.}

The most serious challenge concerns the nature of this research. Unlike ordinary
work on programming languages, the question seems to call for research similar
to the one in the human-computer interaction area. At the same time, a
significant evaluation result demands a good number of cases yielding a serious
amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
around this dilemma is to postulate the equivalent of a {\it homo economicus\/}
for programming languages---the {\em rational programmer\/}---and to implement
this rational programmer as an algorithm. As indicated in the introduction, in
the context of gradual typing a rationally acting programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion.

One derivative challenge to the first is the problem of simulating the
programmer's mistakes. Before a programmer can try to find mistakes, mistakes
have to be made. Technically the problem is the lack of a sufficiently large
available corpus of type errors for a meaningful experiment. Similar to
~\citet{lksfd-popl-2020}, we cover this gap by injecting bugs in typed programs
with mutation analysis~\cite{lipton1971fault, demillo1978hints,
jia2011analysis}. This results into a large number of buggy programs whose
mixed-typed versions are suitable subjects for our experiment. However, mutation
analysis traditionally aims to inject bugs that are sophisticated enough to
challenge extensive test suites. As a consequence, the ill-typed mutants we care
about here are usually deemed \emph{incompetent} and mutation analysis
frameworks are fine-tuned to avoid them. To overcome this problem, we design a
set of mutators specific for type errors inspired by common errors when
programming in Typed Racket.  These type errors cover a wide range of Typed
Racket's type system and moreover cause reliably runtime type errors in
mixed-typed versions of mutants.

Another derivative challenge is the problem of mistakes in type specifications.
Gradual typing enables programmers to move an untyped code base into the typed
realm on an incremental basis. The retroactive addition of a specific type to a
component, even a library, may result from a misunderstanding of the code and
may thus be a mistake itself. Empirical evidence suggests that this scenario is
reasonably common~\cite{sta-nt-base-types, incorrect-ts, wmwz-ecoop-2017}, and
hence must be addressed by any evaluation method. Similarly to the
previous challenge, we deal with this one by designing a set of mutators
that ``break'' type annotations but leave programs otherwise intact.

The second big challenge is the set of seemingly incomparable gradual
typing systems. Despite the fact that the natural, transient and optional
approaches are all well-understood as variants of the same minimal operational model
~\cite{gf-icfp-2018, gfd-oopsla-2019}, their actual implementations live on
different linguistic settings. Hence it is not possible to compare them directly 
on the same programs; code needs to be adjusted in often subtle ways to
transfer from one implementation to another. As a first step to overcome the problem, 
we have implemented a transient and optional variant of Typed Racket.  
However,
implementations on top of the same platform are not sufficient for an
apples-to-apples comparison. As we discuss in detail in
section~\ref{sec:landscape}, the three strategies differ in the kind of
blame information they produce. Natural blames one component while
transient blames a set of components and optional none. To account for these
differences, our competent programmer comes with \emp{modes} that represent 
the alternative ways programmers can interpret blame information. For instance, in
transient, one mode  has the competent programmer pick the oldest element of a blame set 
as the next component to add types. After all, this corresponds to the
earliest point in the program's evaluation that can discover the bug. 
Another mode though, does not distinguish between the elements of the
blame set and instead it requires that the competent programmer types 
all blamed components at once. In order to compare different strategies
across their different modes, we normalize the steps the competent
programmer takes to find the bug by the effort of each step, that is the
number of component it has to type. Finally, besides differences in how
and what they blame, the three gradual typing approaches differ in the type checks they perform. 
To untangle the two, we supplement our experiment with two extra modes. The \emph{vanilla} mode
ignores blame and instead uses information from dynamic type error messages to
select the next component to type. Hence it separates the effect of blame 
from that of the type checks. The \emph{control} mode randomly
selects which part of the program to migrate and serves as the null hypothesis
control for the overall method.

### 2020-07-02 writing/introduction.tex
working on challenges

The paper's second contribution is a set of results from applying the evaluation
method to three distinct blame assignment systems and approximately 20,000
cases\mf{Replace with real number} (see section~\ref{sec:experiment}): Vitousek et
al.'s strategy of tracking typed/untyped boundaries in Reticulated;
Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,
mf-toplas-2009}; and the error messages from failed safety checks in the
underlying programming language.

>>>>

The paper's second contribution is a set of results from applying the evaluation
method to three distinct blame assignment systems and approximately 20,000
cases\mf{Replace with real number} (see section~\ref{sec:experiment}): Vitousek et
al.'s strategy of tracking typed/untyped boundaries in Reticulated;
Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,
mf-toplas-2009}; and the approach of industrial systems that forego blame
and instead rely on error messages from the
underlying programming language.
